An algorithm for the optimization of finite element integration loops

### 2016-12-05 TOMS_Paper/assembly.tex
Change section 7 headings
minor fixes

\subsection{Performance results}
\label{sec:perf-results-forms}

@@

\subsubsection{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$. There are two reasons for this. First, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Second, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

\subsubsection{Autovectorization}

@@

\subsubsection{Mass matrix}

@@

\subsubsection{Helmholtz}

@@

In the case $\langle \mathrm{nf}=0,\ q=2 \rangle$, the 2$\times$ speed-ups achieved by \texttt{cfO2} and \texttt{tens} are due to the remarkable drop in operation count achieved by pre-evaluating reductions: from order $10^4$ (\texttt{ufls} $3 \cdot 10^4$, \texttt{cfO1} $2 \cdot 10^4$) to order $10^3$. By unrolling loops and eliminating operations over zeros, \texttt{tens} eventually executes fewer operations than \texttt{cfO2} (roughly, $10^3$ instead of $2 \cdot 10^3$), but the performance is basically identical because of matrix insertion, which plays again the limiting role. With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous paragraph, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.

Another relevant problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time. This is again caused by the chosen code generation strategies: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure. This choice mitigates the memory transfer (especially between L1 cache and registers) by maximizing register locality.

@@

An interesting experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$. We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$. This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.

\subsubsection{Elasticity}

@@

We now consider the test case $\langle \mathrm{nf}=2,\ q=3 \rangle$ to discuss the relationship between achieved performance and operation count, although analogous considerations apply to the other scenarios. It is difficult to estimate the number of flops executed by the code generated by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler. The calculation is instead simple in all other cases: while \texttt{cfO2} performs roughly $2 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are much closer to $10^6$ operations. In particular, \texttt{ufls} requires 4.3$\times$ more operations than \texttt{cfO2}, whereas the execution time speed-up is only slightly bigger than 2$\times$. The distance between these two values may be attributed to the memory transfer cost -- especially inserting the $3.6 \cdot 10^3$ non-zero entries of the local element matrix -- which, for reasons already discussed, becomes a bottleneck. \texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($84 \cdot 10^4$ instead of $97 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency. Despite requiring roughly $10^5$ fewer operations than \texttt{cfO2}, \texttt{tens} is again affected by the poor low level efficiency of the generated code.

\subsubsection{Hyperelasticity}

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive. For the original code, at the source level $37 \cdot 10^9$ operations are counted. However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have been applied in aggressive way. The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.

Unsurprisingly, the memory transfer impact is less relevant in this hyperelastic than in all previous forms. With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.

>>>>

@@

\subsection{High level view}
Our transformation strategy does not always guarantee minimum execution time. In particular, about 5$\%$ of the test cases (3 out of 56, without counting marginal differences) show that \texttt{cfO2} was not optimal in terms of runtime. The most significant of such test cases is the elastic model with $[q=4,\ \mathrm{nf}=0]$. There are two reasons for this. First, low level optimization can have a significant impact on the actual performance. For example, the aggressive loop unrolling in \texttt{tens} eliminates operations on zeros and reduces the working set size by not storing entire temporaries; on the other hand, preserving the loop structure can maximize the chances of autovectorization. Second, the transformation strategy adopted when $T_H$ is exceeded plays a key role, as we will later elaborate.

\subsection{Autovectorization}

@@

\subsection{Mass matrix}

@@

\subsection{Helmholtz}

@@

In the case $\langle \mathrm{nf}=0,\ q=2 \rangle$, the 2$\times$ speed-ups achieved by \texttt{cfO2} and \texttt{tens} are due to the remarkable drop in operation count achieved by pre-evaluating reductions: from order $10^4$ (\texttt{ufls} $3 \cdot 10^4$, \texttt{cfO1} $2 \cdot 10^4$) to order $10^3$. By unrolling loops and eliminating operations over zeros, \texttt{tens} eventually executes fewer operations than \texttt{cfO2} (roughly, $10^3$ instead of $2 \cdot 10^3$), but the performance is basically identical because of matrix insertion, which plays again the limiting role. With $10^2$ entries to be updated in the global matrix, following the same procedure of the previous section, it is straightforward to verify that $\frac{\text{operation count}}{\text{bytes moved}} \ll 1$, a typical symptom of memory-boundedness.

Another interesting problem instance is $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{cfO2} performs 1.5$\times$ more operations than \texttt{tens}, but still exhibits a better execution time. This is again a consequence of the code generation strategy: while \texttt{tens} prefers to fully unroll the computation to detect all avoidable operations, \texttt{cfO2} preserves the loop structure. This choice mitigates the memory transfer, especially between L1 cache and registers, by maximizing data locality.

@@

A relevant experiment we additionally performed was relaxing the memory threshold by setting $T_H = \operatorname{size}(\mathrm{L3})$. We found that this makes \texttt{cfO2} generally slower for $\mathrm{nf} \geq 2$, with a maximum slow-down of 2.16$\times$ with $\langle \mathrm{nf}=2, q=2\rangle$. This effect could be worse when running in parallel, since the L3 cache is shared and different threads would end up competing for the same resource.

\subsection{Elasticity}

@@

We now consider the test case $\langle \mathrm{nf}=2,\ q=3 \rangle$ to discuss the relationship between achieved performance and operation count, although analogous considerations apply to the other scenarios. It is difficult to estimate the number of flops executed by the code produced by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler. The calculation is instead simple in all other cases: while \texttt{cfO2} performs $2.3 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are close to $10^6$ operations. In particular, \texttt{ufls} requires 4.3$\times$ more operations than \texttt{cfO2}, whereas the execution time speed-up is only slightly bigger than 2$\times$. The distance between these two values may be attributed to the memory transfer cost -- especially inserting the $3.6 \cdot 10^3$ non-zero entries of the local element matrix -- which, for reasons already discussed, becomes a bottleneck. \texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($8.4 \cdot 10^5$ instead of $9.7 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency. Despite requiring roughly $10^5$ fewer operations than \texttt{cfO2}, \texttt{tens} is again affected by the poor low level efficiency of the generated code.

\subsection{Hyperelasticity}

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups were generally massive. In the original code, $37 \cdot 10^9$ operations were counted at the source level. However, the number of operations actually executed is in practice much lower, since by inspection of the assembly code generated by the underlying compiler it is clear that common sub-expressions elimination and loop-invariant code motion are applied in aggressive way. The actual operation count is, however, definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.

Unsurprisingly, the memory transfer impact in the hyperelastic model is less relevant than in all other forms studied. With $\langle \mathrm{nf}=0,\ q=1 \rangle$, \texttt{cfO2} is already 1.4$\times$ faster than the untransformed code, and the gain increases with both $nf$ and $q$.
